---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Katie Knox"
date: "10-2-2025"
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**

- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:

- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
# Load required packages
library(sf)
library(tidyverse)
library(ggplot2)
library(tigris)
library(tidycensus)
library(patchwork)
library(units)
library(knitr)
library(ggspatial)
census_api_key("42bf8a20a3df1def380f330cf7edad0dd5842ce6")

# Load spatial data
pa_counties <- st_read("data/Pennsylvania_County_Boundaries.shp")
hospitals <- st_read("data/hospitals.geojson")
census_tracts <- tracts(state = "PA", cb = TRUE)

census_tracts <- st_transform(census_tracts, st_crs(pa_counties))
hospitals <- st_transform(hospitals, st_crs(pa_counties))

# Check that all data loaded correctly
glimpse(pa_counties)
glimpse(census_tracts)
glimpse(hospitals)

# With ggplot2
p1 <- ggplot(pa_counties) +
  geom_sf() +
  theme_void()

p2 <- ggplot(census_tracts) +
  geom_sf() +
  theme_void()

p3 <- ggplot(hospitals) +
  geom_sf() +
  theme_void()

p1 | p2 | p3

```

**Questions to answer:**

- How many hospitals are in your dataset?
  - 223
- How many census tracts?
  - 3445
- What coordinate reference system is each dataset in?
  - I transformed hospitals and census tracts to PA counties CRS, which is WGS84. Before that, hospitals were also in WGS84, but census tracts were in NAD83. 

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**

- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
# Get demographic data from ACS
pa_tracts_data <- get_acs(
  geography = "tract",
  variables = c(
    median_income = "B19013_001",
    total_pop = "B01003_001",
    over_65 = "B01001_020"  # Population 65 years and over
  ),
  state = "PA",
  year = 2023,
  output = "wide"
)

# Join to tract boundaries
census_tracts <- census_tracts %>%
  left_join(pa_tracts_data, by= "GEOID")

summary(census_tracts)

```

**Questions to answer:**

- What year of ACS data are you using?
  - 2019-2023 5-year ACS data
- How many tracts have missing income data?
  - 62
- What is the median income across all PA census tracts?
  - $70,188


---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:

1. Low median household income (choose an appropriate threshold)
2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}
# Filter for vulnerable tracts based on your criteria
vulnerable_tracts <- census_tracts %>%
  filter(median_incomeE < 42398 | (over_65E/total_popE) > .017)

```

**Questions to answer:**

- What income threshold did you choose and why?
    - I looked at the [2023 poverty guidelines](https://www.pakeys.org/2023-annual-federal-poverty-guidelines/), the average household size in PA in 2023, which was 2.4, and then looked into eligibility for various programs, which is anywhere from 133% FPIG to 215%. I am going with vulnerable household is 215% or less of the 2-person household FPIG, or $42,398.
- What elderly population threshold did you choose and why?
    - In order to control for overall population size, instead of looking at the raw number of elderly, I looked at the over 65 as percent of total population. I defined vulnerable as any tract in the 75th percentile of elderly percent of tract population (1.7% or above).
- How many tracts meet your vulnerability criteria?
    - 1,145
- What percentage of PA census tracts are considered vulnerable by your definition?
    - 33.2% -- about a third. 

---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
#convert to Albers
vulnerable_tracts <- st_transform(vulnerable_tracts, crs = 5070)
hospitals <- st_transform(hospitals, crs = 5070)

# Calculate distance from each tract centroid to nearest hospital
tract_centroids <- st_centroid(vulnerable_tracts)
nearest_hospital <- st_nearest_feature(tract_centroids, hospitals)

vulnerable_tracts <- vulnerable_tracts %>%
  mutate(nearest_hospital_geom = hospitals$geometry[nearest_hospital])

vulnerable_tracts <- vulnerable_tracts %>%
  mutate(distance_to_nearst_hospital = set_units(st_distance(tract_centroids, vulnerable_tracts$nearest_hospital_geom, by_element = TRUE), "mi"))

summary(vulnerable_tracts$distance_to_nearst_hospital) 

```

**Requirements:**

- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection

I chose Albers Conical Equal Area, NAD 83 projection based on [this PA document from 2011's](https://elibrary.dcnr.pa.gov/PDFProvider.ashx?action=PDFStream&docID=1743537&chksum=&revision=0&docName=ndards&nativeExt=pdf&PromptToSave=False&Size=152488&ViewerMode=2&overlay=0) recommendation. 

**Questions to answer:**

- What is the average distance to the nearest hospital for vulnerable tracts?
    - 5.8 miles
- What is the maximum distance?
    - 39.8 miles
- How many vulnerable tracts are more than 15 miles from the nearest hospital?
    - 104

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable

vulnerable_tracts <- vulnerable_tracts %>%
  mutate(underserved = ifelse(as.numeric(distance_to_nearst_hospital) > 15,"Underserved",""))

vulnerable_tracts %>%
  group_by(underserved)%>%
  summarise(n())

```

**Questions to answer:**

- How many tracts are underserved?
    - 104
- What percentage of vulnerable tracts are underserved?
    - 9%
- Does this surprise you? Why or why not?
    - Unfortunately, this does not surprise me, given how rural certain areas of Pennsylvania are and knowing how rural healthcare is a persistent issue in the US in general. However, it would surprise me if some of these tracts are in urban areas. 

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
# Spatial join tracts to counties
pa_counties <- pa_counties %>% st_transform(st_crs(vulnerable_tracts))

vulnerable_per_county <- vulnerable_tracts %>%
  st_join(pa_counties) %>%
  st_drop_geometry()

# Aggregate statistics by county
vulnerable_per_county_stats <- vulnerable_per_county %>%
  group_by(COUNTY_NAM) %>%
  summarise(
    num_vulnerable_tracts = n(),
    num_underserved_tracts = sum(underserved == "Underserved"),
    percent_underserved = sum(underserved == "Underserved")/n(),
    avg_distance_to_nearst_hospital = mean(distance_to_nearst_hospital),
    total_pop = sum(total_popE),
    total_underserved_pop = sum(ifelse(underserved == "Underserved", total_popE, 0))
  )

```

**Required county-level statistics:**

- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population

**Questions to answer:**

- Which 5 counties have the highest percentage of underserved vulnerable tracts?
    - Sullivan
    - Juniata
    - Pike
    - Cameron
    - Snyder
- Which counties have the most vulnerable people living far from hospitals?
    - Clearfield
    - Chester
    - Pike
    - Erie
    - Cumberland
- Are there any patterns in where underserved counties are located?
    - There are a lot of quite rural underserved counties, like Sullivan, Juniata, and Clearfield, but there are also suburban counties of major PA cities like Chester County outside of Philadelphia and Cumberland County outside Harrisburg and Erie County which includes Erie. 

---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
# Create and format priority counties table
vulnerable_per_county_stats %>% 
  arrange(desc(total_underserved_pop)) %>%
  slice_head(n = 10) %>%
  select(
    c("COUNTY_NAM", "total_underserved_pop", "percent_underserved", "avg_distance_to_nearst_hospital")
  ) %>%
  mutate(
    avg_distance_to_nearst_hospital = as.numeric(avg_distance_to_nearst_hospital),
    percent_underserved = paste0(round(percent_underserved*100, 2), "%")
  )%>%
  kable(
    col.names = c("County", "Total Underserved Population", "Percent of County made up of Underserved Tracts", "Average Distance of to Nearest Hospital"),
    digit = 1,
    format.args = list(big.mark = ","),
    align = "l",
    caption = "10 Counties in PA with the Highest Absolute Population more than 15 miles from Nearest Hospital"
  )


```

**Requirements:**

- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map
vulnerable_per_county_stats <- vulnerable_per_county_stats %>%
  left_join(pa_counties, by="COUNTY_NAM")

vulnerable_per_county_stats <- vulnerable_per_county_stats %>%
  st_as_sf() %>% 
  st_transform(st_crs(census_tracts)) %>%
  mutate(percent_underserved = percent_underserved*100
  )

hospitals <- hospitals %>%
  mutate(type = "Hospital")

ggplot(vulnerable_per_county_stats) +
  geom_sf(aes(fill = percent_underserved)) +
  scale_fill_gradient(
    low = "#C3CDFE",    # light periwinkle
    high = "#485EFE",   # dark periwinkle
    name = "Underserved Tracts per County (%)"
  ) +
  geom_sf(
    data = hospitals, 
    #aes(fill = type),       # map fill to type, triggers legend
    size = 3, 
    shape = 21, 
    color = "#FF8600",      # outline color stays fixed
    stroke = 0.8
  ) +
  # Manual fill scale for hospital 'type'
 # scale_fill_manual(
  #  values = c("Hospital" = "#FF8600"),  # orange fill for hospital points
  #  guide = guide_legend(override.aes = list(shape = 21, size = 4, color = "#FF8600")),
  #  name = "Legend"
  #) +
  labs(
    title = "Underserved Tracts per Counties in Pennsylvania",
    subtitle = "Where and underserved tract is one where the center is at least 15 miles from the nearest hospital",
    caption ="Data Sources: "
  ) +
    annotation_north_arrow(             
    location = "tr",                  
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  theme_void() +
    theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 0.5, size = 9, margin = margin(t = 10))
  )

```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map




```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization




```

**Suggested chart types:**
- Histogram or density plot of distances
- Box plot comparing distances across regions
- Bar chart of underserved tracts by county
- Scatter plot of distance vs. vulnerable population size

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/). 

**Note these are just loose suggestions to spark ideas - follow or make your own as the data permits and as your ideas evolve. This analysis should include bringing in your own dataset, ensuring the projection/CRS of your layers align and are appropriate for the analysis (not lat/long or geodetic coordinate systems). The analysis portion should include some combination of spatial and attribute operations to answer a relatively straightforward question**

---

#### Education & Youth Services

**Option A: Educational Desert Analysis**
- **Data:** Schools, Libraries, Recreation Centers, Census tracts (child population)
- **Question:** "Which neighborhoods lack adequate educational infrastructure for children?"
- **Operations:** Buffer schools/libraries (0.5 mile walking distance), identify coverage gaps, overlay with child population density
- **Policy relevance:** School district planning, library placement, after-school program siting

**Option B: School Safety Zones**
- **Data:** Schools, Crime Incidents, Bike Network
- **Question:** "Are school zones safe for walking/biking, or are they crime hotspots?"
- **Operations:** Buffer schools (1000ft safety zone), spatial join with crime incidents, assess bike infrastructure coverage
- **Policy relevance:** Safe Routes to School programs, crossing guard placement

---

#### Environmental Justice

**Option C: Green Space Equity** 
- **Data:** Parks, Street Trees, Census tracts (race/income demographics)
- **Question:** "Do low-income and minority neighborhoods have equitable access to green space?"
- **Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree canopy or park acreage per capita, compare by demographics
- **Policy relevance:** Climate resilience, environmental justice, urban forestry investment
---

#### Public Safety & Justice

**Option D: Crime & Community Resources**
- **Data:** Crime Incidents, Recreation Centers, Libraries, Street Lights
- **Question:** "Are high-crime areas underserved by community resources?"
- **Operations:** Aggregate crime counts to census tracts or neighborhoods, count community resources per area, spatial correlation analysis
- **Policy relevance:** Community investment, violence prevention strategies
---

#### Infrastructure & Services

**Option E: Polling Place Accessibility**
- **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates)
- **Question:** "Are polling places accessible for elderly and disabled voters?"
- **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access
- **Policy relevance:** Voting rights, election infrastructure, ADA compliance

---

#### Health & Wellness

**Option F: Recreation & Population Health**
- **Data:** Recreation Centers, Playgrounds, Parks, Census tracts (demographics)
- **Question:** "Is lack of recreation access associated with vulnerable populations?"
- **Operations:** Calculate recreation facilities per capita by neighborhood, buffer facilities for walking access, overlay with demographic indicators
- **Policy relevance:** Public health investment, recreation programming, obesity prevention

---

#### Emergency Services

**Option G: EMS Response Coverage**
- **Data:** Fire Stations, EMS stations, Population density, High-rise buildings
- **Question:** "Are population-dense areas adequately covered by emergency services?"
- **Operations:** Create service area buffers (5-minute drive = ~2 miles), assess population coverage, identify gaps in high-density areas
- **Policy relevance:** Emergency preparedness, station siting decisions

---

#### Arts & Culture

**Option H: Cultural Asset Distribution**
- **Data:** Public Art, Museums, Historic sites/markers, Neighborhoods
- **Question:** "Do all neighborhoods have equitable access to cultural amenities?"
- **Operations:** Count cultural assets per neighborhood, normalize by population, compare distribution across demographic groups
- **Policy relevance:** Cultural equity, tourism, quality of life, neighborhood identity

---

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/
- Most datasets available as GeoJSON, Shapefile, or CSV with coordinates
- Always check the Metadata for a data dictionary of the fields.

**Additional Sources:**
- **Pennsylvania Open Data:** https://data.pa.gov/
- **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns
- **TIGER/Line (via tigris):** Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with multiple data layers. 
**If you are a beginner, choose something more manageable that helps you understand the basics**


**If you have a different idea:** Propose your own question! Just make sure:
- You can access the spatial data
- You can perform at least 2 spatial operations

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
# Load your additional dataset




```

**Questions to answer:**
- What dataset did you choose and why?
- What is the data source and date?
- How many features does it contain?
- What CRS is it in? Did you need to transform it?

---

2. **Pose a research question**

Write a clear research statement that your analysis will answer.

**Examples:**
- "Do vulnerable tracts have adequate public transit access to hospitals?"
- "Are EMS stations appropriately located near vulnerable populations?"
- "Do areas with low vehicle access have worse hospital access?"

---

3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis










```

**Analysis requirements:**
- Clear code comments explaining each step
- Appropriate CRS transformations
- Summary statistics or counts
- At least one map showing your findings
- Brief interpretation of results (3-5 sentences)

**Your interpretation:**

[Write your findings here]


## Finally - A few comments about your incorporation of feedback!
Take a few moments to clean up your markdown document and then write a line or two or three about how you may have incorporated feedback that you recieved after your first assignment. 

---

## Submission Requirements

**What to submit:**

1. **Rendered HTML document posted to your course portfolio** with all code, outputs, maps, and text
   - Use `embed-resources: true` in YAML so it's a single file
   - All code should run without errors
   - All maps and charts should display correctly

**File naming:** `LastName_FirstName_Assignment2.html` and `LastName_FirstName_Assignment2.qmd`

---

